Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-08-08 isolation_by_coalescence.tex
edited until results

\newcommand{\Dcom}{\widetilde D}

@@

% NOTE: I tend to use semantic linebreaks.
% Please excuse the ragged right margins in the source.

@@

\subsection*{Genetic distance and coalescence time}

\plr{pasted in: divergence is coalescence of lineages}
The genetic distance between
individuals, which provides a good estimate of the average time to a common ancestor along the genome, or
the average coalescence time. Coalescence is the process in which genes or regions of the genome from two
or more individuals, if traced back in time, eventually share a common origin [Wakeley, 2005]. While we
cannot know exactly when this happens, if we assume the rate of mutation per nucleotide per generation is
relatively constant, we can use the number of differences between two copies of a gene to estimate how much
time has passed since their common ancestor. With enough loci, we can calculate an estimate for
the average coalescence time across the genome [Hudson, 2007].

\plr{pasted in: lineages as random walks}
Even though the population dynamics in forwards time are Markov, the dynamics of lineages traced back
in time (the coalescent process) may not be. For example, if fecundity is variable and population density
is low, two lineages near each other are more likely to share a recent common origin, so knowing how one
lineage moves gives you information about how the other lineage is likely to move. This effect becomes
negligible as population density increases [Barton et al., 2002]), so it should not be an issue if the offspring
of any one parent are typically interspersed with the offspring of many others. The methods I develop here
model the coalescent process as Markov, which should be a good approximation for many systems.

\plr{Pasted in, needs transition; make understandable without the below}
The mean commute time of a location to itself is zero,
so to fit to genetic data, we clearly must also account for local diversity.
\plr{say why}
Coalescence time does not have this property,
because coalescence time accounts for both the first meeting time and the eventual coalescence,
while commute time is more analagous to the first meeting time.
For local diversity, we will also have another set of model parameters, $q$, 
where $q_{i}$ approximates the observed coalescence time in location $i$.
As is done by \citet{petkova2016visualizing},
we use commute time to approximate genetic distance as
\begin{align} \label{eq:hcom}
	D_{ij} \approx \Dcom_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
To gain an intuitive feel for why resistance distance is divided by 4, 
consider how, if the movement rates are symmetric, 
a single particle moving from one state to another is similar to (but not the same as)
the first meeting time of two particles, 
one moving forward in time from the first location 
and the second one moving backward in time from the other, 
so it is similar to twice the first meeting time.
Since the commute time is the sum of the hitting times in each direction, 
it is similar to 4 times the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}. 
The average of the local diversity estimates ($q$) are then added 
to the first meeting time estimate to account for the eventual coalescence after meeting.
%\emph{Briefly motivate by describing "first meeting" plus "subsequent coalescence" -> 
%	commute + diversity approximation.}
In this case, these values
are parameters in the model to be inferred instead of within deme coalescence rates ($\gamma$).
Since the values a $q$ are typically fairly easy to infer 
and requiring them all to be the same can significantly decrease the quality of the fit, 
we will not be using an analogue of Problem Modification \ref{probmod:const-coal} 
(where we would only have one parameter for $q$).
Commute time inference has the advantage of only needing to use the base Markov chain
instead of the product chain,
but it also leads to incorrect inferences in certain circumstances.
Using the base Markov chain instead of the product chain is advantageous 
because, as the grid size increases, it is much more computationally efficient 
to work with a Markov chain with $n$ states than $n^2$ states 
since the matrices would be much smaller.

\begin{figure}
\centering
% \includegraphics[scale=.5]{conceptn}
\caption{Illustration of the conceptual differences
betweeen coalescence and commute time between states 1 and 3 
for a continuous time Markov chain with four linearly connected states. 
Coalescence time (left) is the time for two independently moving particles to meet and coalesce 
(it is possible for them to be in the same location without coalescing).
Commute time is time for a single particle starting in one state to travel to another state 
and then back to the original state.
Note how the time axis is scaled to be faster in the figure showing commute time
than the one for coalescence time in order for them to fit in the same space.}
\label{fig:concept_coalcom}
\end{figure}

For a simple example where the coalescence times and commute time approximations are very different, 
consider a system of three islands in a line,
where lineages currently in the outer islands are much more likely to come from the central island
than the other way around.
In this case, lineages are likely to quickly move to the center and coalesce regardless of starting position,
so coalescence time will be relatively short between all individuals.
Commute time between the two outer locations, on the other hand,
will be much longer since it requires the lineage to leave the center.
If we plug real numbers into this example, say, 
reverse time gene flow rate of $1$ into the center and $0.1$ out of it
with $\gamma=1$ everywhere,
then $H_{13}=H_{31}=2.90$ but $\Dcom_{13}=\Dcom_{31}=7.94$.
However, the fact that $\Dcom$ as predicted from the correct parameters may be far from $H$
does not mean that the parameters inferred using resistance distance methods
will necessarily be very wrong.
In principle, it still might be that the parameters for which $\Dcom$ is closest to $H$ 
are close to the real parameters, 
at least in some situations.

\begin{figure}
\centering
% \includegraphics[scale=.5]{three_state_example}
\caption{A simple example where the coalescence time and commute time based approximation
are very different.
In this case, commute time between the two outer locations will be much longer 
than coalescence time.}
\label{fig:3_state}
\end{figure}

@@

First, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the jump rate from $x$ to $y$.
This means that if the Markov chain is at location $X_t$ at time $t$,
then for $x \neq y$, the probability the chain jumps from $x$ to $y$ is proportional to $G_{xy}$,
i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain,
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
By conditioning on $X_\epsilon$,
one can use the Markov property to show that
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
\end{align}
i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$.

@@

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
This can also be seen by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \prod I + I \prod G$,
where $I$ is the identity matrix and $\prod$ is the Kronecker product.

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so can be solved with straightforward linear algebra.
If $\gamma$ is given, then solving for $G$ is an example of a XXX equation \citep{matrixeqn}.
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZH$ is antisymmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $Z$
so that $G + Z$ remains the generator of a valid Markov chain,
but this is not guaranteed in general.

@@

Consider a Markov chain with two states, 
where the rate of movement from state 1 to state 2 is $G_{12}$
and the rate of movement from state 2 to state 1 is $G_{21}$,
so the generator matrix is
\begin{align*}
G = 
    \begin{bmatrix}
        -G_{12}  & G_{12} \\
         G_{21}  & -G_{21}
    \end{bmatrix}.
\end{align*}
Writing out equation \ref{eqn:C_matrix}, we get three unique equations
for the coalescence times:
\begin{align*}
   -1 &= -C_{11} G_{12} + C_{21} G_{12} - C_{11} G_{12} + C_{12} G_{12} - C_{11} \gamma_1 \\
   -1 &=  C_{11} G_{21} - C_{21} G_{21} - C_{21} G_{12} + C_{22} G_{12} \\
   -1 &=  C_{12} G_{21} - C_{22} G_{21} + C_{21} G_{21} - C_{22} G_{21} - C_{22} \gamma_2 .
\end{align*}
Given $C$, we have three equations for the four unknowns, 
$G_{12}$, $G_{21}$, $\gamma_{1}$, and $\gamma_{2}$.
By grouping terms together, factoring back into matrix form, 
and writing $C_{ji}$ instead of $C_{ij}$ when $i > j$,
we have
\begin{align*}
\begin{bmatrix}
-2C_{11} + 2C_{12} & 0                 & -C_{11} & 0       \\
-C_{12} + C_{22}   & C_{11} - C_{12}   & 0       & 0       \\
0                  & 2C_{12} - 2C_{22} & 0       & -C_{22}
\end{bmatrix}
\begin{bmatrix}
G_{12} \\ G_{21} \\ \gamma(1) \\ \gamma(2)
\end{bmatrix}
=
\begin{bmatrix}
-1 \\ -1 \\ -1
\end{bmatrix}.
\end{align*}
This is small enough to be solved symbolically. 
The valid solutions to these equations are those with $G_{12}$, $G_{21}$, $\gamma_1$,
and $\gamma_2$ nonnegative.
If $C_{11} \neq C_{12}$, 
we can write the solution with $\gamma_1$ as the free variable:
\begin{align*}
G_{12} &= \frac{1}{2(C_{11} - C_{12})} - \frac{C_{11}}{2(C_{11} - C_{12})}\gamma_1 \\
G_{21} &= \frac{-2(C_{11} - C_{12}) + (C_{12} - C_{22})}{2(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})}{2(C_{11} - C_{12})^2}\gamma_1 \\
\gamma_2 &= \frac{(C_{11} - 2C_{12} + C_{22})^2}{C_{22}(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})^2}{C_{22}(C_{11} - C_{12})^2}\gamma_1.
\end{align*}
This implies that $\gamma_2$ decreases as $\gamma_1$ increases as long as $C_{12} \neq C_{22}$.
This makes sense because
increasing a rate of coalescence cannot make expected coalescence times longer,
so in order to keep coalescence times the same,
if one coalescence rate is increased, another must be decreased.
This also implies that, in order for $\gamma_2$ to be a value other than $0$, 
we must have $C_{11} - 2C_{12} + C_{22} \neq 0$.
In general, if there is a value or range of values of $\gamma_1$
for which the other movement and coalescence parameters are non-negative,
then a solution exists.

@@

Mean coalescence times estimated from real data are subject to a number of sources of noise
(which we discuss in more detail later).
Here, we use
the exact solutions based on linear algebra developed above
to develop a Bayesian inference method that accounts for noise in the data
and estimates uncertainty in the resulting estimates.
Suppose that $\hat{C}_{ij} = C + \epsilon_{ij}$,
where $\epsilon$ is Normally distributed
with mean $\mathbf{0}$ and covariance $\Sigma$.
For now, we will make a few assumptions about $\epsilon$.
First, that all $\epsilon_{ij}$ are independent, identically distributed $\mathcal{N}(0,\sigma_{\epsilon}^2)$
for $i \leq j$, and $\epsilon_{ij} = \epsilon_{ji}$.
Additionally, we will assume that $C_{ij}$ is large compared to $\sigma_{\epsilon}$
such that the probability that a value of $\hat{C}$ is ever negative is negligible.
Writing $f(\mathbf{x})$ as the probability density function of 
the multivariate normal distribution $\epsilon$,
the likelihood function for the pair $(G,\gamma)$ is
\begin{align*}
\mathcal{L}(G,\gamma;\hat{C}) &= f(\mathcal{C}(G,\gamma) - \hat{C}) \\
			      &= f(-(G^*-\text{diag$(\Gamma)$})^{-1} \mathbf{1} - \hat{C})
\end{align*}
recalling that $\mathcal{C}$ is the hitting time matrix for a given pair $(G,\gamma)$
defined by equation \ref{eq:solve_H}. 
If $\hat{C}$ is a valid coalescence time matrix,
we can find the maximum likelihood estimate (MLE) of $G$;
however, this is in practice unlikely to happen, 
and the likelihood function is still useful 
for providing information about the uncertainty of $G$.

To complete the Bayesian model,
we place priors on the parameters.
Let's say that $i$ is said to be adjacent to $j$ if $i \neq j$ and $G_{ij}$ is one of 
the values known to be non-zero.
Then we could let the prior probability density for the movement parameters be that of an 
exponential random variable with rate parameter $\lambda_{G}$. 
That is, the prior likelihood of $G$ would be
\begin{align*}
\mathcal{L}_G(G) = 
\begin{cases}
	\lambda_{G}^{m_G} \prod_{i \sim j} e^{-\lambda_{G} G_{ij}} 
		\quad & \text{for $G_{ij} \geq 0 \quad \forall \ i \neq j$} \\
	0 & \text{otherwise}
\end{cases}
\end{align*}
where $i \sim j$ means that $i$ and $j$ are adjacent and $m_G$ is the number of positive values of $G$. 
Similarly, we can put a prior density on $\gamma$:
\begin{align*}
\mathcal{L}_{\gamma}(\gamma) = 
\begin{cases}
	\prod_{i}^{n} \lambda_{\gamma} e^{-\lambda_{\gamma}\gamma_i} 
		\quad & \text{for $\gamma_i \geq 0 \quad \forall \ i$} \\
	0 & \text{otherwise.}
\end{cases}
\end{align*}
Recall that $n$ is the number of states in the single particle Markov chain.
If we assume that all $\epsilon_{ij}$ are IID with $\epsilon_{ij} \sim \mathcal{N}(0,\sigma_{\epsilon}^2)$ 
for $i \leq j$
we can write the likelihood for $\hat{C}$:
\begin{align*}
\mathcal{L}(\hat{C};G,\gamma) = \frac{1}{(2 \pi \sigma_{\epsilon}^2)^{m/2}} \prod_{i \leq j}
	\exp\left\{-\frac{(\mathcal{C}_{ij}(G,\gamma) - \hat{C}_{ij})^2}{2 \sigma_{\epsilon}^2}\right\},
\end{align*}
where $m$ is the number of nonmissing values on or above the diagonal in $\hat{C}$.
Then the posterior distribution is
\begin{align} \label{eq:post}
\mathcal{L}(\hat{C};G,\gamma) \mathcal{L}_{G}(G) \mathcal{L}_{\gamma}(\gamma) = 
	\frac{\lambda_{G}^{m_G} \lambda_{\gamma}^n}{(2 \pi \sigma_{\epsilon}^2)^{m/2}}
	\exp\{-\sum_{i \sim j} \lambda_G G_{ij} -\sum_{i}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} (\mathcal{C}_{ij}
		(G,\gamma) - \hat{C}_{ij})^2 \}
\end{align}
with the same support over $G$ and $\gamma$ as in their individual priors.
To sample from this distribution, 
we will use the Metropolis-Hastings algorithm \citep{brooks2011handbook}.

\paragraph{Incomplete observations}
With real world data, 
unless the number of states is small or we have a large amount of data,
our pairwise average coalescent time matrix will have a large number of unobserved values.
For example, if our landscape is a $20 \times 20$ grid and we take $100$ samples, 
each from a different grid square,
we would have only $100(101)/2$ out of $400(401)/2$ of the unique values for $H$. 
(Note that we are including the self-comparisons in this example;
if the organisms were haploid or inbreeding were substantial,
our $H$ matrix would have only $99(100)/2$ usable values.)
To carry out inference in this situation,
we simply use Equation \ref{eq:post},
but the rightmost sum will be only over observed entries of H.

%%%%%%%%%%%%%%%%
\subsection*{Test cases and simulations}

\plr{describe graphs and slim sims}

\plr{MCMC methods: pre-burnin, burnin, number of iterations, trace plots}

>>>>

\newcommand{\comdist}{\widetilde R}

@@

@@

The \emph{commute time} of a Markov chain between locations $i$ and $j$
is the mean time it takes a particle moving according to the rules of the chain, 
started at $i$, to first reach $j$ and then return.
We will denote this by $R_{ij}$, to emphasize that it is also the effective resistance
between $i$ and $j$.
The \emph{coalescence time} is the mean time it takes two particles started at $i$ and $j$
to \emph{coalesce},
which happens at a given rate whenever they are in the same location.
We will denote this by $C_{ij}$.
Why should these be related to genetic distance? 
First, we explain this connection,
and then discuss their computation and application.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Genetic distance and coalescence time}

There are a number of ways to calculate measures of dissimilarity using genetic data \citep{distances}.
In this paper we use mean pairwise divergence:
for two genomes, it is calculated as the density of sites at which the two differ.
The genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
from the two genomes back to their most recent common ancestor at that site;
under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
pairwise divergence is therefore an unbiased estimate of the mean time to most recent common ancestor,
averaged across the genome \citep{hudson2007variance,ralphXX}.
It is natural to model this process by following the two lineages back through time
(and across space)
until they find their most recent common ancestor,
i.e., until they coalesce.
It is common to model the geographic distribution of a species
as a collection of randomly mating populations
that exchange occasional migrants,
thus discretizing geographic space.
If the genome is selectively netural,
then the motion of these lineages back through time
forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
In this framework, each lineage performs a random walk across the populations,
with movement probabilities depending on the flux of individuals between populations.
This provides the link between genetic distance and random walks.

For data, we are given genotypes of individuals and their geographic coordinates;
then, we divide these individuals into groups,
and compute the genetic distance between group $i$ and group $j$
as the mean divergence between a pair of genomes, chosen one from each group.
We denote the resulting genetic distance $D_{ij}$;
since divergence is symmetric, $D_{ij} = D_{ji}$.
We compute $D_{ii}$ by sampling without replacement from group $i$,
so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
As discussed below, we will find commute times so that the mean commute time between
$i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
so we must account for local diversity.
We do this in the same way as \citet{petkova2016visualizing},
by introducing another set of model parameters, $q$, 
that fill the same role as the coalescence rates, $\gamma$.
Then, we use commute time to approximate genetic distance as
\begin{align} \label{eq:commute_approx}
	D_{ij} \approx \comdist{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
The factor of four appears to make commute time more similar to coalescence time:
the commute time covers the path from $i$ to $j$ twice
(once in both directions),
while the coalescence time covers the distance only once,
and in half the time, because two particles are moving.
Furthermore, coalescence time includes within-location diversity,
because particles do not coalesce instantly --
commute time is most analogous to the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}.

\begin{figure}
\centering
    % \includegraphics[scale=.5]{conceptn}
    \caption{
    Illustration of the conceptual differences
    betweeen coalescence and commute time between states 1 and 3 
    for a continuous time Markov chain with four linearly connected states. 
    Coalescence time (left) is the time for two independently moving particles to meet and coalesce 
    (it is possible for them to be in the same location without coalescing).
    Commute time is time for a single particle starting in one state to travel to another state 
    and then back to the original state.
    Note how the time axis is scaled to be faster in the figure showing commute time
    than the one for coalescence time in order for them to fit in the same space.
    } \label{fig:concept_coalcom}
\end{figure}

@@

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the jump rate from $x$ to $y$.
This means that if the Markov chain is at location $X_t$ at time $t$,
then for $x \neq y$, the probability the chain jumps from $x$ to $y$ is proportional to $G_{xy}$,
i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain,
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
By conditioning on $X_\epsilon$,
one can use the Markov property to show that
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
\end{align}
i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$.

@@

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
This can also be seen by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$,
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra.
If $\gamma$ is given, then solving for $G$ is an example of a XXX equation \citep{matrixeqn}.
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZH$ is antisymmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $Z$
so that $G + Z$ remains the generator of a valid Markov chain,
but this is not guaranteed in general.

@@

Consider a Markov chain with two states, 
where the rate of movement from state 1 to state 2 is $G_{12}$
and the rate of movement from state 2 to state 1 is $G_{21}$,
so the generator matrix is
\begin{align*}
G = 
    \begin{bmatrix}
        -G_{12}  & G_{12} \\
         G_{21}  & -G_{21}
    \end{bmatrix}.
\end{align*}
Writing out equation \ref{eqn:C_matrix}, we get three unique equations
for the coalescence times:
\plr{simplify these equations and remove matrix version?}
\begin{align*}
   -C_{11} G_{12} + C_{21} G_{12} - C_{11} G_{12} + C_{12} G_{12} - C_{11} \gamma_1 &= -1 \\
    C_{11} G_{21} - C_{21} G_{21} - C_{21} G_{12} + C_{22} G_{12}                   &= -1 \\
    C_{12} G_{21} - C_{22} G_{21} + C_{21} G_{21} - C_{22} G_{21} - C_{22} \gamma_2 &= -1 .
\end{align*}
Given $C$, we have three equations for the four unknowns, 
$G_{12}$, $G_{21}$, $\gamma_{1}$, and $\gamma_{2}$.
By grouping terms together, factoring back into matrix form, 
and writing $C_{ji}$ instead of $C_{ij}$ when $i > j$,
we have
\begin{align*}
\begin{bmatrix}
-2C_{11} + 2C_{12} & 0                 & -C_{11} & 0       \\
-C_{12} + C_{22}   & C_{11} - C_{12}   & 0       & 0       \\
0                  & 2C_{12} - 2C_{22} & 0       & -C_{22}
\end{bmatrix}
\begin{bmatrix}
G_{12} \\ G_{21} \\ \gamma(1) \\ \gamma(2)
\end{bmatrix}
=
\begin{bmatrix}
-1 \\ -1 \\ -1
\end{bmatrix}.
\end{align*}
This is small enough to be solved symbolically. 
The valid solutions to these equations are those with $G_{12}$, $G_{21}$, $\gamma_1$,
and $\gamma_2$ nonnegative.
If $C_{11} \neq C_{12}$, 
we can write the solution with $\gamma_1$ as the free variable:
\begin{align*}
G_{12} &= \frac{1}{2(C_{11} - C_{12})} - \frac{C_{11}}{2(C_{11} - C_{12})}\gamma_1 \\
G_{21} &= \frac{-2(C_{11} - C_{12}) + (C_{12} - C_{22})}{2(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})}{2(C_{11} - C_{12})^2}\gamma_1 \\
\gamma_2 &= \frac{(C_{11} - 2C_{12} + C_{22})^2}{C_{22}(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})^2}{C_{22}(C_{11} - C_{12})^2}\gamma_1.
\end{align*}
This implies that $\gamma_2$ decreases as $\gamma_1$ increases as long as $C_{12} \neq C_{22}$.
This makes sense because
increasing a rate of coalescence cannot make expected coalescence times longer,
so in order to keep coalescence times the same,
if one coalescence rate is increased, another must be decreased.
This also implies that, in order for $\gamma_2$ to be a value other than $0$, 
we must have $C_{11} - 2C_{12} + C_{22} \neq 0$.
In general, if there is a value or range of values of $\gamma_1$
for which the other movement and coalescence parameters are non-negative,
then a solution exists.

@@

Mean coalescence times estimated from real data are subject to a number of sources of noise
(which we discuss in more detail later).
Here, we use
the exact solutions based on linear algebra developed above
to develop a Bayesian inference method that accounts for noise in the data
and estimates uncertainty in the resulting estimates.
To do this, we model genetic distances as equal to coalescence times plus noise:
$D_{ij} = C_{ij} + \epsilon_{ij}$,
where $\epsilon_{ij}$ are independent and Normally distributed
with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
except that $\epsilon_{ij} = \epsilon_{ji}$.
(We imagine that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
so that resulting values are still positive.)
Using the calculations above, 
given $G$ and $\gamma$ we can compute the corresponding coalescence times
-- call these $\mathcal{C}(G, \gamma)$ --
so our model is that $D$ has a Gaussian distribution 
with mean $\mathcal{C}(G, \gamma)$.
The maximum likelihood estimate of $G$ could be found directly;
however, this in practice will quite likely contain negative movement rates,
and does not allow us to impose constraints
(such as only allowing a subset of movement rates to be nonzero).

Therefore, we use Bayesian inference,
placing independent exponential priors on the parameters --
on those movement rates $G_{ij}$ that are not constrained to be zero,
and on the coalescence rates, $\gamma$.
Concretely, suppose that the spatial arrangement of populations is given by a graph,
and only movement rates corresponding to edges in the graph are allowed to be nonzero.
Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
that there are $m_G$ edges in the graph,
and that each of the $n$ populations has its own coalescence rate.
Then, the resulting log-posterior density is
\begin{align} \label{eq:post}
\mathcal{L}(D, G, \gamma) 
    &=
    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
	- \sum_{i \sim j} \lambda_G G_{ij}  \\
    &\qquad
    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
        \right)^2 
\end{align}
Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
respectively.

Inference with commute times is done in exactly the same way,
except that $q$ replaces $\gamma$,
and instead of $\mathcal{C}(G,\gamma)$ we use $\comdist(G, q)$
as defined in equation \eqref{eq:commute_approx}.

It is not required in this approach to have samples from every spatial location,
allowing inference on incompletely sampled spatial discretizations.
To carry out inference in this situation,
we simply use Equation \ref{eq:post},
but only summing over observed entries of $D_{ij}$.

\paragraph{MCMC methods}
To sample from the posterior distribution of equation \eqref{eq:post}, 
we used R \citep{R} 
to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
Starting locations were chosen by sampling from the prior distribution.
Before the usual ``burn-in'' phase before sampling for inference begins,
we included a period of ``pre-burn-in'' which used the same MCMC procedure
with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
portion of parameter space.
Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
and $3 \time 10^6$ iterations of burn-in,
followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions.
We used standard methods to assess convergence and mixing.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Test cases and simulations}

We compared inference under coalescence and commute time models
using both data ``from the model'' and more realistic simulations.
For the first category, we designed a number of graphs to test particular aspects of inference,
with migration rates specified on each directed edge
(plotted below using igraph \citep{igraph}).
Then we calculated exact coalescence and commute times on these graphs,
and added Gaussian noise independently to each entry of the resulting matrices
(while preserving symmetry)
as described above.
Particular parameters and levels of noise are given in the Results.

These simulations are expected to be nearly equivalent to population-based simulations
(either forwards-time or coalescent);
the only difference being that noise terms about the theoretical mean should be somewhat correlated.

\paragraph{Forwards-time simulations}
To produce realistic data, we implemented forwards-time simulations using SLiM v3.0 \citep{haller2017},
with individuals living across continuous, two-dimensional space (sometimes with barriers),
from which we recorded genomic data.
We performed three types of simulation:
uniform migration on a continuous square landscape, 
uniform migration on a rectangular landscape with barriers to migration, 
and biased migration on a square landscape.
Each simulation had $10^4$ diploid, hermaphroditic individuals with nonoverlapping generations.
Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
truncated at $3 \sigma_d$.
Offspring are dispersed away from the mother using the same kernel,
conditioned to lie within the habitable area.
In each simulation, $\sigma_d$ was chosen 
so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
we also model local competition. 
To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
between each pair of nearby individuals;
then, the fitness of each individual is set to $\exp(-x/40)$, 
where $x$ is the sum of their interaction strengths with other individuals.
This reduces fitness for individuals in denser areas.
Individuals are chosen to reproduce proportional to their fitness.
Clumpiness is similar to what is seen in 
Figure \ref{fig:ind_locs_5x3b_1}

Individuals have 10 pairs of ``chromosomes'' with $2 \times 10^5$ loci each 
and a recombination rate of $10^{-8}$ per generation per locus.
Mutation rate was $10^{-8}$ per locus for the simulation with uniform migration and no barrier,
and $10^{-7}$ for the remaining simulations.
In all cases, the number of mutations was more than adequate to reliably estimate genetic distance.

To compute genetic distances,
geographic space was partitioned into rectangular regions as depicted below.
within each region, 
individuals are sampled uniformly for ``genotyping'' 
from among those within the middle three quarters of each dimension of rectangle,
as shown for one situation in Figure \ref{fig:ind_locs_5x3b_1}.
This protocol was chosen as a compromise between sampling all individuals 
from near the center of each location,
which could result in a large number of close relatives being sampled, 
leading to an underestimate of the mean genetic distance for individuals in that location
and uniform sampling from the whole area of each grid, 
which may result in many sampled individuals being very close to other grid squares.
Mean genetic distances for each pair of regions is then computed as described above
from genome sequence provided by SLiM.
To estimate uncertainty, we computed the standard error of these mean distances
using the minumum number of individuals in each location.
